An overview on temporal databases.\\ New approaches and proposals.

### 2012-01-23 trunk/book/sources/fuzzyTemporalDatabases.tex
update with the bipolar querying of temporal databases. some typos corrected.

%
% Fuzzy temporal databases
%
Some software applications deal with uncertainty related to time. e.g. a logistics company which transports packages from one destination to another. The time when the package would be at the end point may be stimated but not precisely known. These applications need to manage the uncertainty with respect to the temporal attributes of the objects stored in the database.

@@

\subsection{Querying fuzzy temporal databases}
\label{subsec:querying-time-intervals}
One of the main pourposes of a database is to allow the information retrieval. The standard query language for databases is SQL ~\cite{Mel93}, however, there are several proposals to extend the SQL language for transaction-time databases~\cite{Sarda90}, valid-time databases~\cite{gad92} and bi-temporal ~\cite{TSQL}. Some authors have studied how to support temporal querying in SQL~\cite{Snodgrass98}.

@@

\begin{definition}
\textbf{(query specification)}
A query $\tilde{Q}$ is given by:
\begin{equation}
\tilde{Q} = \left( Q^{time}, Q \right)
\end{equation}
Here $Q$ are the possibly fuzzy non-temporal contraints and $Q^{time}$ are the temporal constraints specified in the query:
\end{definition}

@@

\subsubsection{Query evaluation}
Query satisfaction in a (fuzzy) relational database is a matter of degree. Tipically, each constraint provided by the user is evaluated and results in a \emph{satisfaction degree} $s \in \left[ 0,1,\right]$. The value 0 denotes a total dissatisfaction while the value 1 denotes a total satisfaction.

The query $\tilde{Q}$ is evaluated in two parts that happen simultaneously:
\begin{itemize}
\item
The non-temporal part, $Q$ is evaluated resulting in a satisfaction degree noted as $e_Q(r)$ for each record $r$.
\item
The temporal part, $Q^{time}$ is evaluated depending on $AR$. A specific set of ill-known constraints is considered depending on the Allen relation in $AR$. Then both possibility and necessity measures are computed and aggregated.
\end{itemize}

\subsubsection{Ranking and aggregation}
For each temporal attribute $r_{time}$ for the record $r$, both possibility $\Pos \left( Q_{time}\right)$ and necessity $\Nec \left( Q_{time} \right)$ measures are obtained.

\subsubsection{Bipolarity in temporal database querying}

>>>>

%
% Fuzzy temporal databases
%
Some software applications deal with uncertainty related to time. e.g. a logistics company which transports packages from one destination to another. The time when the package would be at the end point may be estimated but not precisely known. These applications need to manage the uncertainty with respect to the temporal attributes of the objects stored in the database.

@@

\subsection{Querying fuzzy temporal databases}
\label{subsec:querying-time-intervals}
One of the main purposes of a database is to allow the information retrieval. The standard query language for databases is SQL ~\cite{Mel93}, however, there are several proposals to extend the SQL language for transaction-time databases~\cite{Sarda90}, valid-time databases~\cite{gad92} and bi-temporal ~\cite{TSQL}. Some authors have studied how to support temporal querying in SQL~\cite{Snodgrass98}.

@@

\begin{definition}
\textbf{(query specification)}
A query $\tilde{Q}$ is given by:
\begin{equation}
\label{eq:query-specification}
\tilde{Q} = \left( Q^{time}, Q \right)
\end{equation}
Here $Q = \left \lbrace q_1, \cdots, q_n \right \rbrace$ are the possibly fuzzy non-temporal constraints and $Q^{time}$ are the temporal constraints specified in the query:
\end{definition}

%%specify better this:
%\begin{svgraybox}
%\begin{definition}
%\label{def:query-constraint}
%\textbf{(query constraint)}
%$Q$ is the 
%A query constraint $q_a$ for an attribute $a$ with attribute domain $\mathcal{D}$, is a restriction in the subset of the values in the attribute domain:
%\begin{equation}
%\label{eq:query-constraint}
%q_a = \left \lbrace d \mid d \subseteq \mathcal{D} \right \rbrace
%\end{equation}
%\end{definition}

@@

\subsubsection{Query evaluation}
\label{subsubsec:query-evaluation}
Query satisfaction in a (fuzzy) relational database is a matter of degree. Typically, each constraint provided by the user is evaluated and results in a \emph{satisfaction degree} $s \in \left[ 0,1,\right]$. The value 0 denotes a total dissatisfaction while the value 1 denotes a total satisfaction.

\begin{definition}
\label{def:evaluation-function}
\textbf{(evaluation function)}
An evaluation function $e_{Q} \left( r \right)$ is a mapping from the attribute value from the row $r$ in the database with respect to the constraints $Q = \left \lbrace q_1, \cdots, q_n \right \rbrace$ to the unit interval $\left[ 0,1 \right]$. 
\begin{equation}
\label{eq:evaluation-function} 
e_{Q} = t \mid t \in \left[ 0,1 \right]
\end{equation}
\end{definition}

Each element in $\tilde{Q}$ is evaluated independently:
\begin{itemize}
\item
The non-temporal part, $Q$ is evaluated resulting in a satisfaction degree noted as $e_Q(r)$ for each record $r$.
\item
The temporal part, $Q^{time}$ is evaluated depending on $AR$. A specific set of ill-known constraints is considered depending on the Allen relation in $AR$. Then both possibility and necessity measures are computed and aggregated.
\end{itemize}

\subsubsection{Ranking and aggregation}
For each temporal attribute $r_{time}$ for the record $r$, both possibility $\Pos \left( Q_{time}\right)$ and necessity $\Nec \left( Q_{time} \right)$ measures are obtained. Then, the measures are combined in order to obtain an evaluation score $e_{Q_{time} \left( r \right)}$. The evaluation function is computed by:

\begin{equation}
\label{eq:temporal-eval}
e_{Q_{time} \left( r \right)} = \left \lbrace t \mid t = \left( \frac{\Pos \left( Q_{time}\right) + \Nec \left( Q_{time} \right)}{2} \right)   \in \left[ 0,1\right] \right \rbrace
\end{equation}

This measure provides a natural score because of the following property of the necessity measure: 
\begin{equation}
\label{eq:necessity} %iff or iif?
\Nec \left( Q_{time} \right) > 0  \Longleftrightarrow  \Pos \left( Q_{time}\right) = 1
\end{equation}

The final ranking is given by a convex combination of both temporal and non-temporal evaluations:

\begin{equation}
\label{eq:convex-combination}
e_{final} \left( r \right) = \left \lbrace t \mid t = \omega \ast e_{Q} \left( r \right) + \left( 1- \omega \right) \ast e_{Q^{time}}, t,\omega \in \left[0,1\right] \right \rbrace
\end{equation}

\subsubsection{Bipolarity in temporal database querying}
\label{subsubsec:bipolarity}
Humans express their preferences in both positive and negative statements. This property is interesting in querying of databases, because sometimes, the user does not know exactly his / her preferences but what he / she does not want. This introduces the necessity of a bipolar way of querying databases. Therefore, the bipolar querying of databases is a querying technique which allows to express both positive and negative criteria in the database query.

In this subsection we will address a proposal for the bipolar querying of temporal databases made by Billiet~\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. First of all, the query structure is shown. Then, both evaluation and ranking proposed by the authors are explained. The model deals with a fuzzy valid-time specification as in~\cite{garrido2009}.

\paragraph{Query structure}
The approach in this work follows the query specification given in \eqref{eq:query-specification}. The query has a global time demand, $Q^{time}$ and here, the regular query constraints specified by $Q$ are split into positive and negative preferences:

\begin{equation}
\label{eq:bipolar-specification}
\tilde{Q} = \left \lbrace Q^{time}, \left( Q^{pos}, Q^{neg} \right) \right \rbrace
\end{equation}

Here $Q^{pos}$ and $Q^{neg}$ represent the positive and the negative criteria, respectively. The temporal demand in $Q^{time}$ may be specified as in ~\cite{garrido2009} or as in ~ \cite{Pon11}, see section \ref{subsec:representation-time-intervals}.
.
\paragraph{Query evaluation}
As explained previously in the query evaluation section (\ref{subsubsec:query-evaluation}), each element in $\tilde{Q}$ is evaluated independently:

\begin{itemize}
\item
The query $Q$ has now two elements: $Q^{pos}$ and $Q^{neg}$. Each element is again evaluated independently and the result is a tuple $\left(s,d \right)$ in which $s$ is the satisfaction degree for $Q^{pos}$. $d$ is called dissatisfaction degree for $Q^{neg}$. The tuple $\left(s,d \right) s,d \in \left[0,1 \right]$ is called \emph{Bipolar Satisfaction Degree}.
\item
The temporal specification is evaluated as explained in \ref{subsubsec:query-evaluation} and the result is a value in the unit interval.
\end{itemize}

\paragraph{Ranking}
In order to present the results to the user, it is necessary to design a combination function. This function allows the classification of the results in the unit interval the preferred way is to made a convex combination as  presented in equation \eqref{eq:convex-combination}.

### 2012-01-23 trunk/book/sources/intro.tex
update with the bipolar querying of temporal databases. some typos corrected.

%temporal reasoning and fuzzy temporal reasoning.
Next to the manage of temporal information is the temporal reasoning~\cite{Allen83}. Dubois and Prade ~\cite{Dubois:jucs_9_9:fuzziness_and_uncertainty_in},\cite{Dubois89} deal with fuzzyness and uncertainty in temporal reasoning, but this topic is outside the scope of this work.

@@

In Section \ref{sec:time-domain} an study about the time and its properties is done. Section \ref{sec:temporal-databases} is an overview of the main problems when managing the time in a database, the different proposals for solving them and the temporal database proposals. The section concludes with a comparison among the different proposals. In Section \ref{sec:fuzzy-temporal-databases} we present a sum up with the proposals for handling the imprecision in temporal databases and the main deficiences we have detected. Finally, there are several commercial temporal database management systems  \emph{TDMBS} like \cite{oracle2009},\cite{posgree2009},\cite{teradata2011},\cite{timedb2005}. All of them are analyzed and compared later on this work.

>>>>

%temporal reasoning and fuzzy temporal reasoning.
Next to the manage of temporal information is the temporal reasoning~\cite{Allen83}. Dubois and Prade ~\cite{Dubois:jucs_9_9:fuzziness_and_uncertainty_in},\cite{Dubois89} deal with fuzziness and uncertainty in temporal reasoning, but this topic is outside the scope of this work.

@@

In Section \ref{sec:time-domain} an study about the time and its properties is done. Section \ref{sec:temporal-databases} is an overview of the main problems when managing the time in a database, the different proposals for solving them and the temporal database proposals. The section concludes with a comparison among the different proposals. In Section \ref{sec:fuzzy-temporal-databases} we present a sum up with the proposals for handling the imprecision in temporal databases and the main deficiencies we have detected. Finally, there are several commercial temporal database management systems  \emph{TDMBS} like \cite{oracle2009},\cite{posgree2009},\cite{teradata2011},\cite{timedb2005}. All of them are analyzed and compared later on this work.

### 2012-01-23 trunk/book/sources/prelim.tex
update with the bipolar querying of temporal databases. some typos corrected.

\subsection{\label{subsec:poss}Possibilistic variables}
Informally, a theory of uncertainty considers a set of uncertain events and tries to quantify the confidence about the occurence of those events. Consider therefore a set of outcomes $\Omega$ and let each subset of $\Omega$ be called an \emph{event}. Hereby, $\Omega$ is called the \emph{certain event} and can thus be regarded as a point of reference. Similarly, $\emptyset$ is called the \emph{impossible event}. The quantification of confidence in a theory of uncertainty is achieved by means of a \emph{confidence measure}~\cite{Choquet53}.
\begin{definition}
\label{def:conf-measure}
Consider a universe of outcomes $\Omega$. A confidence measure is defined by a function:
\begin{equation}
g:\Pow(\Omega)\rightarrow[0,1]
\end{equation}
that satisfies:
\begin{eqnarray}
g\left(\emptyset\right)&=&0\\
g\left(\Omega\right)&=&1\\
A\subseteq B&\Rightarrow&g(A)\leq g(B)
\end{eqnarray}
and where $\Pow(\Omega)$ represents the power set of $\Omega$.
\end{definition}
From the constraints in Definition \ref{def:conf-measure}, it can be seen that $\emptyset$ and $\Omega$ are indeed respectively the impossible and the certain event. In addition, the confidence that is assigned to events is monotonic. Due to this monotonicity, we have that:
\begin{eqnarray}
\forall A\in\Omega:\forall B\in\Omega&:&g\left(A\cup B\right)\geq\max\left(g\left(A\right),g\left(B\right)\right)\\
\forall A\in\Omega:\forall B\in\Omega&:&g\left(A\cap B\right)\leq\min\left(g\left(A\right),g\left(B\right)\right).
\end{eqnarray}
We can thus derive two extreme measures of confidence and these measures play a crucial role in the theory of possibility~\cite{Dubois88}. Consider a confidence measure $\Pi$ that satisfies the following constraint:
\begin{equation}
\label{EQ:MAXITIVITY}
\forall A\subseteq\Omega:\forall B\subseteq\Omega:\Pi\left(A\cup B\right) = \max\left(\Pi(A),\Pi(B)\right).
\end{equation}
The function $\Pi$ is then called a \emph{possibility measure} and due to \eqref{EQ:MAXITIVITY}, it is said that $\Pi$ is \emph{maxitive}~\cite{Dubois88}. For each event $A\subseteq\Omega$, $\Pi(A)$ expresses the possibility that $A$ occurs. Due to the maxitivity of $\Pi$ and under the constraint that $\Omega$ is finite, we can easily extract information about the possibility of separate outcomes in the form of a \emph{possibility distribution} $\pi$ defined by:
\begin{equation}
\pi:\Omega\rightarrow[0,1]:\pi(u)=\Pi\left(\left\{u\right\}\right).
\end{equation}
Equivalently, we have that:
\begin{equation}
\forall A\subseteq\Omega:\Pi(A)=\sup_{u\in A}\pi (u).
\end{equation}
A possibility distribution is normalized in the sense that:
\begin{equation}
\sup_{u\in\Omega}\pi(u)=1.
\end{equation}
In order to ensure the existence of a possibility distribution when $\Omega$ is infinite, the maxitivity constraint \eqref{EQ:MAXITIVITY} should be replaced by:
\begin{equation}
\Pi\left(\bigcup_{i}A_i\right)=\sup_{i}\Pi\left(A_i\right)
\end{equation}
for any family $\left\{A_i\right\}$ of indexed subsets of $U$~\cite{Nguyen79}.

>>>>

\subsection{\label{subsec:poss}Possibilistic variables}
Informally, a theory of uncertainty considers a set of uncertain events and tries to quantify the confidence about the occurrence of those events. Consider therefore a set of outcomes $\Omega$ and let each subset of $\Omega$ be called an \emph{event}. Hereby, $\Omega$ is called the \emph{certain event} and can thus be regarded as a point of reference. Similarly, $\emptyset$ is called the \emph{impossible event}. The quantification of confidence in a theory of uncertainty is achieved by means of a \emph{confidence measure}~\cite{Choquet53}.
\begin{definition}
\label{def:conf-measure}
Consider a universe of outcomes $\Omega$. A confidence measure is defined by a function:
\begin{equation}
g:\Pow(\Omega)\rightarrow[0,1]
\end{equation}
that satisfies:
\begin{eqnarray}
g\left(\emptyset\right)&=&0\\
g\left(\Omega\right)&=&1\\
A\subseteq B&\Rightarrow&g(A)\leq g(B)
\end{eqnarray}
and where $\Pow(\Omega)$ represents the power set of $\Omega$.
\end{definition}
From the constraints in Definition \ref{def:conf-measure}, it can be seen that $\emptyset$ and $\Omega$ are indeed respectively the impossible and the certain event. In addition, the confidence that is assigned to events is monotonic. Due to this monotonicity, we have that:
\begin{eqnarray}
\forall A\in\Omega:\forall B\in\Omega&:&g\left(A\cup B\right)\geq\max\left(g\left(A\right),g\left(B\right)\right)\\
\forall A\in\Omega:\forall B\in\Omega&:&g\left(A\cap B\right)\leq\min\left(g\left(A\right),g\left(B\right)\right).
\end{eqnarray}
We can thus derive two extreme measures of confidence and these measures play a crucial role in the theory of possibility~\cite{Dubois88}. Consider a confidence measure $\Pi$ that satisfies the following constraint:
\begin{equation}
\label{EQ:MAXITIVITY}
\forall A\subseteq\Omega:\forall B\subseteq\Omega:\Pi\left(A\cup B\right) = \max\left(\Pi(A),\Pi(B)\right).
\end{equation}
The function $\Pi$ is then called a \emph{possibility measure} and due to \eqref{EQ:MAXITIVITY}, it is said that $\Pi$ is \emph{maxitive}~\cite{Dubois88}. For each event $A\subseteq\Omega$, $\Pi(A)$ expresses the possibility that $A$ occurs. Due to the maxitivity of $\Pi$ and under the constraint that $\Omega$ is finite, we can easily extract information about the possibility of separate outcomes in the form of a \emph{possibility distribution} $\pi$ defined by:
\begin{equation}
\pi:\Omega\rightarrow[0,1]:\pi(u)=\Pi\left(\left\{u\right\}\right).
\end{equation}
Equivalently, we have that:
\begin{equation}
\forall A\subseteq\Omega:\Pi(A)=\sup_{u\in A}\pi (u).
\end{equation}
A possibility distribution is normalized in the sense that:
\begin{equation}
\sup_{u\in\Omega}\pi(u)=1.
\end{equation}
In order to ensure the existence of a possibility distribution when $\Omega$ is infinite, the maxitivity constraint \eqref{EQ:MAXITIVITY} should be replaced by:
\begin{equation}
\Pi\left(\bigcup_{i}A_i\right)=\sup_{i}\Pi\left(A_i\right)
\end{equation}
for any family $\left\{A_i\right\}$ of indexed subsets of $U$~\cite{Nguyen79}.

### 2012-01-23 trunk/book/sources/timeDomain.tex
update with the bipolar querying of temporal databases. some typos corrected.

Several operators are defined in order to compare temporal elements. Allen~\cite{Allen83} first described the relations between time intervals and to a lesser extent, between time points. Figure \ref{fig:allen} shows the temporal relations. There are some proposals that can be applied to both crisp and fuzzy temporal interval ~\cite{ohlbach2004},\cite{nagypal2003},\cite{schockaert08}. For this fuzzy comparisons~\cite{garrido2009} proposed some different temporal relations that can be easyly implemented by means of regular fuzzy operations.

>>>>

Several operators are defined in order to compare temporal elements. Allen~\cite{Allen83} first described the relations between time intervals and to a lesser extent, between time points. Figure \ref{fig:allen} shows the temporal relations. There are some proposals that can be applied to both crisp and fuzzy temporal interval ~\cite{ohlbach2004},\cite{nagypal2003},\cite{schockaert08}. For this fuzzy comparisons~\cite{garrido2009} proposed some different temporal relations that can be easily implemented by means of regular fuzzy operations.
